\section{Introduction}
\label{sec:intro}

Texture partitioning is not just semantic segmentation with unusual labels. The target regions are often defined by repeated local structure, appearance statistics, and transition behavior rather than by named object categories. The classical natural and synthetic texture datasets in this literature make that distinction explicit, and they also expose the practical bottleneck: texture boundaries can be weak, disconnected, or globally ambiguous even when local evidence is strong~\cite{khan2015stld,khan2017c2f,khan2018learnedstld,mikes2022texturebenchmark}. A model built primarily for broad semantic segmentation is therefore not automatically expected to behave as a texture segmenter.

That is exactly the puzzle raised by SAM-style foundation models. SAM and SAM2 were not trained as texture segmenters~\cite{kirillov2023sam,ravi2024sam2}, yet recent adaptation work such as TextureSAM shows that much stronger texture behavior can be induced by fine-tuning on those classical natural and synthetic benchmarks~\cite{cohen2025texturesam,cohen2026texturesam}. This creates a more interesting question than ``can fine-tuning help?'' Fine-tuning working at all suggests that useful texture-relevant structure may already exist inside the frozen model. The scientific question is therefore whether frozen SAM already contains texture understanding, and if so, where that understanding lives and how much of it can be extracted without retraining the backbone or proposal generator.

We study that question through the lens of \emph{recoverability}. A frozen SAM-style model exposes at least two possible loci of texture-relevant evidence. One route explores feature space: coarse frozen features may already organize the target partition before any learned decoder or proposal reasoning. The other explores proposal masks: the generated mask bank may already contain the right fragments, even when the default output fails to assemble them into a coherent texture partition. These routes are complementary rather than hierarchical. The feature route asks what the frozen representation reveals; the proposal-/mask-space route asks what the generated masks reveal. Taken together, they separate latent texture organization from recoverable texture partitions.

\input{figures/fig_recoverability_loci}

Empirically, the paper keeps the historical natural/synthetic benchmark pair as gold-standard anchors, but it does not stop there. Those anchors remain valuable precisely because they are canonical and difficult, yet as historical benchmarks they are not comprehensive on their own: they do not by themselves span broader public natural-image texture crops or more naturalistic synthetic seams. We therefore broaden the natural side with a curated ADE20K textural-crop subset assembled from public refined crops~\cite{detextureade20k2026} and the synthetic side with a ControlNet-stitched bridge dataset~\cite{zhang2023controlnet}. A separate CAID shoreline benchmark is retained in the appendix as a real-world complement with exact masks and unusually large contiguous partitions in remote-sensing / overhead imagery~\cite{wang2025caid}.

\paragraph{Research questions.} In this paper we ask four questions.
\begin{enumerate}[leftmargin=1.4em]
\item \textbf{RQ1:} Does frozen SAM contain recoverable texture understanding at all?
\item \textbf{RQ2:} What does the frozen feature space reveal about that texture understanding?
\item \textbf{RQ3:} What do the generated masks / proposal masks reveal about that texture understanding?
\item \textbf{RQ4:} Under what regimes do the two routes differ in what they reveal?
\end{enumerate}

\paragraph{Answers and contributions.} The paper's contributions are best read as answers to those questions.
\begin{itemize}[leftmargin=1.2em]
\item To answer \textbf{RQ1} and \textbf{RQ2}, we introduce a feature-space exploration over frozen multiscale features. A lightweight clustering probe, qualitative gallery, and scale / positionality diagnostics show that coarse frozen features already organize nontrivial texture structure.
\item To answer \textbf{RQ1} and \textbf{RQ3}, we introduce a proposal-/mask-space exploration over the generated mask bank. A lightweight learned readout above frozen proposals recovers strong texture partitions under matched evaluation on the historical natural and synthetic anchors, improving coherence from 0.6163 to 0.7013 ARI on the natural anchor while keeping mIoU comparable, and improving the synthetic anchor from 0.5140/0.7526 to 0.7195/0.7791.
\item To test whether that answer is confined to the older anchor pair, we add two broader complements as a benchmark contribution in their own right: a curated ADE20K textural-crop subset on the natural side, and a ControlNet-stitched bridge dataset on the synthetic side. The same frozen-bank readout improves the matched ADE20K crop subset from 0.4715/0.3306 to 0.5043/0.3762 and the matched bridge route from 0.6436/0.5519 to 0.6803/0.6039.
\item To answer \textbf{RQ4}, we compare what the two routes reveal jointly. Selector controls, oracle decompositions, and the historical natural-versus-synthetic contrast show when the generated mask bank already contains a coherent singleton answer and when fragmented mask evidence still requires structured commitment.
\item Across these questions, the paper keeps its claim narrow. We do not argue that frozen SAM fully solves texture segmentation. We argue that frozen SAM contains texture-relevant structure in both frozen features and generated masks, and that the two routes reveal complementary aspects of that structure.
\end{itemize}

\paragraph{Roadmap.} Section~\ref{sec:problem} formalizes recoverability and defines the two loci of frozen evidence. Section~\ref{sec:feature} addresses \textbf{RQ2} through feature-space exploration. Section~\ref{sec:proposal} addresses \textbf{RQ3} through proposal-/mask-space exploration. Section~\ref{sec:data} defines the historical benchmark anchors and the broader natural/synthetic complements. Section~\ref{sec:results} brings the two routes together to answer \textbf{RQ1} and \textbf{RQ4} through matched anchor-benchmark evidence, broadened complement routes, learned-selector controls, oracle analysis, and residual failure diagnostics. Section~\ref{sec:discussion} closes the loop on what the paper now knows about frozen SAM's texture understanding and what remains unresolved.
