\section{Recoverability Formulation}
\label{sec:problem}

Let a frozen SAM-style segmenter expose two kinds of evidence for an image $x$: a multiscale feature representation $\phi(x)$ and a proposal bank $\mathcal{P}(x)=\{m_i\}_{i=1}^{N_x}$. The downstream target is a binary texture partition $Y \in \{0,1\}^{H\times W}$. Depending on the benchmark, the foreground side may be canonical (STLD) or label-symmetric (RWTD and the appendix routes), but the output is always one binary partition.

We define \emph{recoverability} as the performance of a readout that must keep the underlying segmenter frozen. The readout may inspect $\phi(x)$, $\mathcal{P}(x)$, or both, but it may not retrain the backbone or alter the proposal generator. This yields two loci:
\begin{align}
\hat{Y}_{\mathrm{feat}} &= R_{\mathrm{feat}}(\phi(x)), \\
\hat{Y}_{\mathrm{prop}} &= R_{\mathrm{prop}}(x, \mathcal{P}(x)).
\end{align}
The first readout asks what texture structure is already latent in frozen features. The second asks what texture structure is already present in the generated mask bank and whether fragmented mask evidence can be assembled into a coherent task partition.

For the proposal route, the frozen bank is first encoded proposal-wise as $d_i = E(x, m_i)$. A learned pair model scores potentially adjacent proposals,
\begin{align}
s(i,j) = f_{\mathrm{pair}}(\psi(d_i, d_j, m_i, m_j)),
\end{align}
and connected components induced by scores above a threshold $\tau$ define candidate merged regions $\mathcal{C}(x)$. A learned component scorer then ranks those candidates,
\begin{align}
q(C) = f_{\mathrm{comp}}(g(C; x, \mathcal{P}(x))), \qquad
C^\star(x) = \arg\max_{C \in \mathcal{C}(x)} q(C).
\end{align}
RWTD adds an optional denser rescue bank $\mathcal{P}^{+}(x)$ and a learned switch that decides whether some rescue candidate $R_k \in \mathcal{P}^{+}(x)$ should replace the conservative core $C^\star(x)$.

\paragraph{Representation-level recoverability.} The feature route uses the lightest possible readout. It clusters coarse frozen \texttt{backbone\_fpn} features, optionally after pooling or flip-averaging, and then scores the resulting binary partition. Its role is to answer \textbf{RQ2}: if such a simple probe already succeeds at all, then the backbone is not texture-blind. Because the available feature artifacts come from an archived probe protocol rather than the official matched evaluator used by the proposal route, we interpret this route as evidence of latent structure rather than mix it into the matched proposal-route tables.

\paragraph{Proposal-level recoverability.} The proposal route treats the frozen bank as incomplete evidence. The readout must decide which fragments are mutually compatible, which merged component should be trusted as the conservative core, and, on RWTD only, when a denser completion is worth the risk. This route answers \textbf{RQ3}: if a lightweight learned readout can recover a strong final partition from frozen proposals, then texture failure in the default output does not imply texture absence in the generated masks.

\paragraph{Fair comparison.} Recoverability claims are only meaningful when evaluator and subset conventions are fixed within each block. The main paper therefore keeps the two routes in their own protocol-aligned comparisons: the feature probe is reported under its archived probe protocol, while proposal-route tables use matched RWTD and STLD comparison blocks only. Unmatched numbers are not mixed across routes.

\input{tables/route_protocol}
