\begin{abstract}
Does frozen SAM contain texture understanding even though it was built mainly for semantic segmentation? Recent fine-tuning results suggest that texture-relevant structure may already exist inside the frozen model. We investigate that possibility through two complementary routes. Feature-space exploration asks what is already latent in frozen multiscale features; proposal-/mask-space exploration asks what is already recoverable from the generated mask bank without retraining the backbone or proposal generator. We study both routes on historical natural and synthetic texture datasets, then broaden each side with a benchmark complement of its own: the natural side with a visually selected textural crop subset from the public DeTexture ADE20K gallery, and the synthetic side with a ControlNet-stitched bridge dataset that preserves exact labels while making transitions more naturalistic. The feature route reveals that coarse frozen features already support plausible texture partitions and that positional debiasing such as flip-averaging materially affects what those features reveal. The proposal route shows that generated masks also contain substantial recoverable texture structure: on the classical natural benchmark it improves coherence from 0.6163 to 0.7013 ARI against a public TextureSAM rerun while keeping mIoU comparable, and on the classical synthetic benchmark it improves from 0.5140/0.7526 to 0.7195/0.7791. The ADE20K and ControlNet complements keep the same reading, improving the matched ADE20K crop subset from 0.4715/0.3306 to 0.5043/0.3762 and the matched bridge route from 0.6436/0.5519 to 0.6803/0.6039 under their invariant evaluators. Together, the two routes indicate that frozen SAM contains texture-relevant structure in both representation and generated-mask space, while the broadened benchmark set shows that this recoverability story is not confined to the historical natural/synthetic pair alone. \codelinksentence
\end{abstract}
